---
title: "Untitled"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
            warnings = FALSE,
            message = FALSE,
            comment = "#>",
            #results = "hide",
            digits = 4,
            error = FALSE)
## clean the R environment
graphics.off()
rm(list = ls())
freshr::freshr()
## load packages
library(here, quietly = TRUE)
library(rio, quietly = TRUE)
library(skimr, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(gtsummary, quietly = TRUE)
library(flextable, quietly = TRUE)
## check the directory for the file
# here::dr_here()
here::set_here()
## the figure or results should be saved 
# paste0("foldername/Sfilename_workingresult_", 
#      Sys.Date(), ".filetype")
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 300px;
  max-width: 1000px;
  overflow-y: auto;
  background-color: inherit;
}
```

```{r}
?i_am
i_am("01_week1.qmd")
here()
skim(hptn.lec1)
```

# Department of Biostatistics and Bioinformatics

- Lee Kennedy-Shaffer, PhD Assistant Professor, Department of Biostatistics


- Bhramar Mukherjee, PhD Senior Associate Dean of Public Health Data Science and Data Equity; Anna M.R. Lauder Professor of Biostatistics; Professor of Epidemiology (Chronic Diseases) and of Statistics and Data Science


- Amy Justice, MD, PhD CNH Long Professor of Medicine and Public Health, Yale University


- Reza Yaesoubi, PhD (moving to UCSF seemingly)


- Emma Zang, Associate Professor of Sociology, Biostatistics, and Global Affairs








